\bf What is the Best Way to Display Two Time Series to Read Association?

### 2016-01-20 paper/paper.Rnw
done with intro

\title{\bf Comparing the Power of Plot Designs to Reveal Correlation}

@@

There have been a few studies related to plotting temporal data.  \citet{Robbins} writes a blog and consults on plotting financial data. She is critical of the overlaid line graph, because it is misleading if the purpose is to read the difference between the two series. Her arguments date back a long time to \citet{playfair,tufte,friendly}. It is well-established that we cannot read differences between two series accurately, that the difference should be plotted separately if that is the information of interest. \citet{Vanderplas} examines a related problem, dubbed the ``sine illusion'', where the human eye is is fooled into erroroneously thinking that variation is larger at the peaks and troughs of seasonal time series. \citet{Vanderplas} developed a mathematical correction to the illusion.

@@

There has been other research on the perception of temporal displays, such as in \citet{Javed}, but they do not examine perception of association. Javed et al.  examined optimal ways to visualise local features, such as, which variable had the highest value at a given point in time, and global features, such as, comparing the size of the overall slope of many different variables. The finding was that a different graphical layout is optimal for each type of task. However, they did not test for the perception of correlation, which can be treated as both a local and a global feature within the graph. The slope of each variable must be compared at a point in time, and they must have a persistent relationship across at least the majority of the series. However there is a quandary, the visual features that improve local tasks are unsuited to global tasks and vice versa; so we must now find a form that is well suited to assess both types of tasks simultaneously. This paper addresses this deficiency. Section 2 describes the lineup method utilised to rigorously compare plot designs. Section 3 explains the experimental design.  Section 4 contains the results of the experiment and Section 5 discusses the implications of the findings.

Visual Inference background next

\section{The Lineup Protocol}

@@

\section{Experimental Design}

@@

\subsection{Detection, Time and Confidence}

@@

\subsection{Selection Analysis}

@@

\subsection{Modelling Graphical Power}

@@

\subsection{Modelling Time Taken}

>>>>

\title{\bf What is the Best Way to Display Two Time Series to Read Association?}

@@

There have been a few studies related to plotting temporal data.  \citet{Robbins} writes a blog and consults on plotting financial data. She is critical of the overlaid line graph, because it is misleading if the purpose is to read the difference between the two series. Her arguments date back a long time to Pfari, as reported in \citet{Tufte:1986} and \citet{friendly}. It is well-established that we cannot read differences between two series accurately, that the difference should be plotted separately if that is the information of interest. \citet{Vanderplas} examines a related problem, dubbed the ``sine illusion'', where the human eye is is fooled into erroroneously thinking that variation is larger at the peaks and troughs of seasonal time series. \citet{Vanderplas} developed a mathematical correction to the illusion.  \citet{Javed} examined the perception of many different sorts of features in multiple time series, and found that there was no optimal plot design for reading all of the different features tested. They did not examine perception of association between two series, though.

@@

%There has been other research on the perception of temporal displays, such as in \citet{Javed}, but they do not examine perception of association. Javed et al.  examined optimal ways to visualise local features, such as, which variable had the highest value at a given point in time, and global features, such as, comparing the size of the overall slope of many different variables. The finding was that a different graphical layout is optimal for each type of task. However, they did not test for the perception of correlation, which can be treated as both a local and a global feature within the graph. The slope of each variable must be compared at a point in time, and they must have a persistent relationship across at least the majority of the series. However there is a quandary, the visual features that improve local tasks are unsuited to global tasks and vice versa; so we must now find a form that is well suited to assess both types of tasks simultaneously.

This paper rigorously examines the perception of association between two time series in two different displays, and with an additional inquiry into whether smoothing of series affects the perception, using the lineup protocol. Section 2 describes the lineup protocol, and the background to its use for comparing the power of plot designs for different purposes. Section 3 explains the experimental design.  Section 4 contains the results of the experiment and Section 5 discusses the implications of the findings.

\section{The lineup protocol for comparing plot designs}

@@

\section{Experimental design}

@@

\subsection{Detection, time and confidence}

@@

\subsection{Selection analysis}

@@

\subsection{Modelling graphical power}

@@

\subsection{Modelling time taken}
